---
title: "Solution Exercice 1 - Introduction to Optimal Design with PFIM 7.1"
author: "Lucie Fayette"
date: "2026-06-07"
output: html_document
---
# Exercice 1

This exercise illustrates:

- Definition of a 1-compartment PK model
- Evaluation of an experimental design using the Fisher Information Matrix (FIM)
- Comparison of designs
- Optimization of sampling times using Simplex and Multiplicative algorithm

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = TRUE,
  message = TRUE,
  error = TRUE  
)
library(PFIM)
library(ggplot2)
library(dplyr)
# library(gridExtra)
library(tibble)
```


# Q1 evaluation of an empirical design


## 1. MODEL DEFINITION
```{r}
modelEquations = list(  "RespPK" = "dose_RespPK/V * ka/(ka - Cl/V) * (exp(-Cl/V * t) - exp(-ka * t))")
```


```{r}
modelFromLibrary = list("PKModel" = "Linear1FirstOrderSingleDose_kaClV")

```


## 2. POPULATION PARAMETERS
```{r}
# model parameters
modelParameters = list(
  ModelParameter( name = "V",    distribution = LogNormal( mu = 8, omega = sqrt(0.02) )),
  ModelParameter( name = "Cl",   distribution = LogNormal( mu = 0.13, omega = sqrt(0.06) ) ),
  ModelParameter( name = "ka",   distribution = LogNormal( mu = 1.6, omega = sqrt(0.70) ) ) )
```

## 3. ERROR MODEL
```{r}
errorModelRespPK = Combined1( output = "RespPK", sigmaInter = 0.6, sigmaSlope = 0.07 )
modelError = list( errorModelRespPK )
```

## 4. DESIGN DEFINITION
```{r}
## administration 
administrationRespPK = Administration( outcome = "RespPK", timeDose = c( 0 ), dose = c( 100 ) )

## sampling times
samplingTimesRespPK = SamplingTimes( outcome = "RespPK", samplings = c( 0.5, 1, 2, 6, 9, 12, 24, 36, 48, 72, 96, 120 ) )

## arms definition
empiricalArm = Arm( name = "empiricalArm",
                    size = 32,
                    administrations = list( administrationRespPK ) ,
                    samplingTimes   = list( samplingTimesRespPK ) )

empiricalDesign = Design( name = "empiricalDesign", arms = list( empiricalArm ) )
```


## 5. FISHER INFORMATION MATRIX COMPUTATION
```{r}
evaluationEmpiricalDesignFIMSettings = Evaluation( name = "",
                                                   modelEquations = modelEquations,
                                                   # modelFromLibrary = modelFromLibrary,
                                                   
                                                   modelParameters = modelParameters,
                                                   modelError = modelError,
                                                   designs = list( empiricalDesign ),
                                                   outputs = list( "RespPK" ),
                                                   fimType = 'population' )

evaluationEmpiricalDesignFIM = run( evaluationEmpiricalDesignFIMSettings )
show( evaluationEmpiricalDesignFIM )
```

# Q2 Evalaution of a two-arm design

```{r}
samplingsArm1 = SamplingTimes( outcome = "RespPK", samplings = c( 0.5, 2, 9, 24, 48, 96 ) )
samplingsArm2 = SamplingTimes( outcome = "RespPK", samplings = c( 1, 6, 12, 36, 72, 120 ) )

## arms definition
Arm1 = Arm( name = "Arm1",
            size = 16,
            administrations = list( administrationRespPK ) ,
            samplingTimes   = list( samplingsArm1 ) )
Arm2 = Arm( name = "Arm2",
            size = 16,
            administrations = list( administrationRespPK ) ,
            samplingTimes   = list( samplingsArm2 ) )

designTwoArms = Design( name = "designTwoArms", arms = list( Arm1, Arm2) )
```

```{r}
evaluationTwoArmsDesignFIMSettings = Evaluation( name = "",
                                                 modelEquations = modelEquations,
                                                 modelParameters = modelParameters,
                                                 modelError = modelError,
                                                 designs = list( designTwoArms ),
                                                 outputs = list( "RespPK" ),
                                                 fimType = 'population' )

evaluationTwoArmsDesignFIM = run( evaluationTwoArmsDesignFIMSettings )
show( evaluationTwoArmsDesignFIM )
```

```{r}
# Extract D-criterion for design comparison
getDcriterion(evaluationEmpiricalDesignFIM)
getDcriterion(evaluationTwoArmsDesignFIM)

# Extract RSE for design comparison
RSE_EmpiricalDesign = getRSE(evaluationEmpiricalDesignFIM)
RSE_TwoArmsDesign = getRSE(evaluationTwoArmsDesignFIM)

RSE_EmpiricalDesign <- tibble::rownames_to_column(RSE_EmpiricalDesign, "Parameter")
RSE_TwoArmsDesign <- tibble::rownames_to_column(RSE_TwoArmsDesign, "Parameter")


RSE_comparison = RSE_EmpiricalDesign%>% 
  mutate(RSE_EmpiricalDesign = RSE) %>% 
  left_join(RSE_TwoArmsDesign %>% 
              rename(RSE_TwoArmsDesign = RSE))
RSE_comparison
```


```{r}
ggplot( RSE_EmpiricalDesign%>% 
          mutate(Design = "RSE_EmpiricalDesign" ) %>% 
          rbind(RSE_TwoArmsDesign %>% 
                  mutate(Design = "RSE_TwoArmsDesign" )),
        aes(x = Parameter, y = RSE, fill = as.factor(Design))) +
  geom_bar( stat="identity", position="dodge") +
  ylab("RSE(%)") + 
  labs(fill = "Design") + 
  theme_bw()
```



# Q3 PK sample optimisation with Simplex algorithm

```{r}
# sampling times
samplingsArmOptimisation = SamplingTimes( outcome = "RespPK", samplings = c( 0.5, 2, 24, 48, 96) )

# sampling times constraints
samplingConstraintsSimplexRespPK  = SamplingTimeConstraints( outcome = "RespPK",
                                                             initialSamplings = c( 0.5, 2, 24, 48, 96 ),
                                                             samplingsWindows = list( c(1/6,120) ),
                                                             numberOfTimesByWindows = c(5),
                                                             minSampling = c(1/6) )

armWithSimplexConstraints = Arm( name = "armWithSimplexConstraints",
                                 size = 32,
                                 administrations = list( administrationRespPK ),
                                 samplingTimes   = list( samplingsArmOptimisation ),
                                 samplingTimesConstraints = list( samplingConstraintsSimplexRespPK ) )


designWithSimplexConstraints = Design( name = "designWithSimplexConstraints", arms = list( armWithSimplexConstraints ), numberOfArms = 32 )

```


```{r}
# RUN SIMPLEX OPTIMISATION
optimizationSimplexSettings = Optimization( name = "Optimisation Simplex",
                                            modelEquations = modelEquations,
                                            modelParameters = modelParameters,
                                            modelError = modelError,
                                            optimizer = "SimplexAlgorithm",
                                            optimizerParameters = list( pctInitialSimplexBuilding = 20,
                                                                        maxIteration = 500,
                                                                        tolerance = 1e-6,
                                                                        showProcess = TRUE  ),
                                            designs = list( designWithSimplexConstraints ),
                                            fimType = "population",
                                            outputs = list( "RespPK" ) )

optimizationSimplexFIM = run( optimizationSimplexSettings )
show(optimizationSimplexFIM)

```


# Q4 PK sample optimisation with Multiplicative algorithm

```{r}

administrationConstraintsMultRespPK = AdministrationConstraints( outcome = "RespPK", doses = list( 100 ) )


samplingConstraintsMultRespPK  = SamplingTimeConstraints( outcome = "RespPK",
                                                          initialSamplings = c( 0.5, 1, 2, 6, 12, 24, 48, 72, 96, 120 ),
                                                          numberOfsamplingsOptimisable = 5,
                                                          # fixedTimes = c()
)


armWithMultConstraints = Arm( name = "armWithMultConstraints",
                              size = 32,
                              administrations = list( administrationRespPK ),
                              administrationsConstraints = list( administrationConstraintsMultRespPK ),
                              
                              samplingTimes   = list( samplingsArmOptimisation ),
                              samplingTimesConstraints = list( samplingConstraintsMultRespPK ) )


designWithMultConstraints = Design( name = "designWithMultConstraints", arms = list( armWithMultConstraints ), numberOfArms = 32 )

```


```{r}
# RUN MULTIPLICATIVE ALGORITHM OPTIMISATION
optimizationMultSettings = Optimization( name = "Optimisation Mult",
                                         modelEquations = modelEquations,
                                         modelParameters = modelParameters,
                                         modelError = modelError,
                                         optimizer = "MultiplicativeAlgorithm",
                                         optimizerParameters = list( lambda = 0.99,
                                                                     numberOfIterations = 1000,
                                                                     weightThreshold = 0.001,
                                                                     delta = 1e-04, showProcess = T ),
                                         designs = list( designWithMultConstraints ),
                                         fimType = "population",
                                         outputs = list( "RespPK" ) )

optimizationMultFIM = run( optimizationMultSettings )
show(optimizationMultFIM)
```


# Bonus 
```{r}

# administrationConstraintsMultRespPK2 = AdministrationConstraints( outcome = "RespPK", doses = list( 100, 200 ) )

samplingConstraintsMultRespPK2  = SamplingTimeConstraints( outcome = "RespPK",
                                                           initialSamplings = c( 0.5, 1, 2, 6, 12, 24, 48, 72, 96, 120 ),
                                                           numberOfsamplingsOptimisable = 6,
                                                           # fixedTimes = c()
)

armWithMultConstraints2 = Arm( name = "armWithMultConstraints",
                               size = 32,
                               administrations = list( administrationRespPK ),
                               administrationsConstraints = list( administrationConstraintsMultRespPK ),
                               
                               samplingTimes   = list( samplingsArmOptimisation ),
                               samplingTimesConstraints = list( samplingConstraintsMultRespPK2 ) )


designWithMultConstraints2 = Design( name = "designWithMultConstraints", arms = list( armWithMultConstraints2 ), numberOfArms = 32 )


optimizationMultSettings2 = Optimization( name = "Optimisation Mult",
                                          modelEquations = modelEquations,
                                          modelParameters = modelParameters,
                                          modelError = modelError,
                                          optimizer = "MultiplicativeAlgorithm",
                                          optimizerParameters = list( lambda = 0.99,
                                                                      numberOfIterations = 1000,
                                                                      weightThreshold = 0.001,
                                                                      delta = 1e-04, showProcess = T ),
                                          designs = list( designWithMultConstraints2 ),
                                          fimType = "population",
                                          outputs = list( "RespPK" ) )

optimizationMultFIM2 = run( optimizationMultSettings2 )
show(optimizationMultFIM2)
```





