---
title: "Solution Exercice 2 - Introduction to Optimal Design with PFIM 7.1"
author: "Lucie FAYETTE"
date: "2026-06-07"
output: html_document
---

# Exercice 2

This exercise illustrates:

- Definition of a 1-compartment PK model including a discrete covariate
- Evaluation of an experimental design using the Fisher Information Matrix (FIM)
- Comparison of designs
- Optimization of sampling times
- Extension to a PK/PD model

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = TRUE,
  message = TRUE,
  error = TRUE  
)
# devtools::load_all("D:/PMX/PROJECTS/LUCIE/CesuPopPKPD/PFIM_7_1_beta/PFIM")
library(PFIM)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(tibble)
```


# Q1 evaluation of an empirical design

## 1. MODEL DEFINITION
```{r}
modelEquations = list(  "RespPK" = "dose_RespPK/V * ka/(ka - Cl/V) * (exp(-Cl/V * t) - exp(-ka * t))")
```


## 2. POPULATION PARAMETERS
```{r}
# model parameters
modelParameters = list(
  ModelParameter( name = "V",    distribution = LogNormal( mu = 8, omega = sqrt(0.02) )),
  ModelParameter( name = "Cl",   distribution = LogNormal( mu = 0.13, omega = sqrt(0.06) ) ),
  ModelParameter( name = "ka",   distribution = LogNormal( mu = 1.6, omega = sqrt(0.70) ) ) )
```

## 3. ERROR MODEL
```{r}
errorModelRespPK = Combined1( output = "RespPK", sigmaInter = 0.6, sigmaSlope = 0.07 )
modelError = list( errorModelRespPK )
```

## 4. DESIGN DEFINITION
```{r}
## administration 
administrationRespPK = Administration( outcome = "RespPK", timeDose = c( 0 ), dose = c( 100 ) )

## sampling times
samplingTimesRespPK = SamplingTimes( outcome = "RespPK", samplings = c( 0.5, 1, 2, 6, 9, 12, 24, 36, 48, 72, 96, 120 ) )

## arms definition
empiricalArm = Arm( name = "empiricalArm",
                    size = 32,
                    administrations = list( administrationRespPK ) ,
                    samplingTimes   = list( samplingTimesRespPK ) )

empiricalDesign = Design( name = "empiricalDesign", arms = list( empiricalArm ) )
```

## 5. COVARIATE DEFINITION
```{r}
# Fixed covariate: CYP2C9 on Cl
CYP2C9 = Covariate(
  name = "CYP2C9", categories = c("Wild","Others"),
  categoriesProportions = c(0.6, 0.4),
  effects = list( "Others" = c( "Cl" = log(0.5) ) )
)
```

## 6. FISHER INFORMATION MATRIX COMPUTATION
```{r}
evaluationEmpiricalDesignFIMSettings = Evaluation( name = "",
                                                   modelEquations = modelEquations,
                                                   modelParameters = modelParameters,
                                                   modelError = modelError,
                                                   modelCovariates = list( CYP2C9 ), 
                                                   modelCovariatesEquation = "exponential",
                                                   designs = list( empiricalDesign ),
                                                   outputs = list( "RespPK" ),
                                                   fimType = 'population' )

evaluationEmpiricalDesignFIM = run( evaluationEmpiricalDesignFIMSettings )
show( evaluationEmpiricalDesignFIM )
```

```{r}
testsCovariate = covariateTest(evaluationEmpiricalDesignFIM,
                               thetaL       = log( 0.80 ),
                               thetaU       = log( 1.25 ),
                               target_power = 0.80,
                               alpha        = 0.05 )

show( testsCovariate )
```


# Q2 Evalaution of a two-arm design

```{r}
samplingsArm1 = SamplingTimes( outcome = "RespPK", samplings = c( 0.5, 2, 9, 24, 48, 96 ) )
samplingsArm2 = SamplingTimes( outcome = "RespPK", samplings = c( 1, 6, 12, 36, 72, 120 ) )

## arms definition
Arm1 = Arm( name = "Arm1",
            size = 16,
            administrations = list( administrationRespPK ) ,
            samplingTimes   = list( samplingsArm1 ) )
Arm2 = Arm( name = "Arm2",
            size = 16,
            administrations = list( administrationRespPK ) ,
            samplingTimes   = list( samplingsArm2 ) )

designTwoArms = Design( name = "designTwoArms", arms = list( Arm1, Arm2) )
```

```{r}
evaluationTwoArmsDesignFIMSettings = Evaluation( name = "",
                                                 modelEquations = modelEquations,
                                                 modelParameters = modelParameters,
                                                 modelError = modelError,
                                                 modelCovariates = list( CYP2C9 ), 
                                                 modelCovariatesEquation = "exponential",
                                                 designs = list( designTwoArms ),
                                                 outputs = list( "RespPK" ),
                                                 fimType = 'population' )

evaluationTwoArmsDesignFIM = run( evaluationTwoArmsDesignFIMSettings )
show( evaluationTwoArmsDesignFIM )
```

```{r}
# Extract D-criterion for design comparison
getDcriterion(evaluationEmpiricalDesignFIM)
getDcriterion(evaluationTwoArmsDesignFIM)

# Extract RSE for design comparison
RSE_EmpiricalDesign = getRSE(evaluationEmpiricalDesignFIM)
RSE_TwoArmsDesign = getRSE(evaluationTwoArmsDesignFIM)

RSE_EmpiricalDesign <- tibble::rownames_to_column(RSE_EmpiricalDesign, "Parameter")
RSE_TwoArmsDesign <- tibble::rownames_to_column(RSE_TwoArmsDesign, "Parameter")


RSE_comparison = RSE_EmpiricalDesign%>% 
  mutate(RSE_EmpiricalDesign = RSE) %>% 
  left_join(RSE_TwoArmsDesign %>% 
              rename(RSE_TwoArmsDesign = RSE))
RSE_comparison
```

```{r}
ggplot( RSE_EmpiricalDesign%>% 
  mutate(Design = "RSE_EmpiricalDesign" ) %>% 
  rbind(RSE_TwoArmsDesign %>% 
              mutate(Design = "RSE_TwoArmsDesign" )),
  aes(x = Parameter, y = RSE, fill = as.factor(Design))) +
  geom_bar( stat="identity", position="dodge") +
  ylab("RSE(%)") + 
  labs(fill = "Design") + 
  theme_bw()
```


```{r}
testsCovariate = covariateTest(evaluationTwoArmsDesignFIM,
                               thetaL       = log( 0.80 ),
                               thetaU       = log( 1.25 ),
                               target_power = 0.80,
                               alpha        = 0.05 )

show( testsCovariate )

testsCovariate@significance
```


# Q3 PK sample optimisation

```{r}
# sampling times
samplingsArmOptimisation = SamplingTimes( outcome = "RespPK", samplings = c( 0.5, 2, 24, 48, 96) )


administrationConstraintsMultRespPK = AdministrationConstraints( outcome = "RespPK", doses = list( 100 ) )


samplingConstraintsMultRespPK  = SamplingTimeConstraints( outcome = "RespPK",
                                                          initialSamplings = c( 0.5, 1, 2, 6, 12, 24, 48, 72, 96, 120 ),
                                                          numberOfsamplingsOptimisable = 5,
                                                          # fixedTimes = c()
)

```


```{r}
armWithMultConstraints = Arm( name = "armWithMultConstraints",
                              size = 32,
                              administrations = list( administrationRespPK ),
                              administrationsConstraints = list( administrationConstraintsMultRespPK ),
                              
                              samplingTimes   = list( samplingsArmOptimisation ),
                              samplingTimesConstraints = list( samplingConstraintsMultRespPK ) )


designWithMultConstraints = Design( name = "designWithMultConstraints", arms = list( armWithMultConstraints ), numberOfArms = 32 )


```


```{r}
# RUN MULTIPLICATIVE ALGORITHM OPTIMISATION
optimizationMultSettings = Optimization( name = "Optimisation Mult",
                                         modelEquations = modelEquations,
                                         modelParameters = modelParameters,
                                         modelError = modelError,
                                         modelCovariates = list( CYP2C9 ), 
                                         modelCovariatesEquation = "exponential",
                                         optimizer = "MultiplicativeAlgorithm",
                                         optimizerParameters = list( lambda = 0.99,
                                                                     numberOfIterations = 1000,
                                                                     weightThreshold = 0.001,
                                                                     delta = 1e-04, showProcess = T ),
                                         designs = list( designWithMultConstraints ),
                                         fimType = "population",
                                         outputs = list( "RespPK" ) )

optimizationMultFIM = run( optimizationMultSettings )
show(optimizationMultFIM)
```



# Q4 extension to a PKPD model
```{r}
# 1. MODEL DEFINITION
modelEquationsPKPD = list(  "Deriv_RespPK" = "dose_RespPK/V * ka  * exp( -ka * t ) - Cl/V * RespPK",
                            "Deriv_RespPD" = "Rin*(1-Imax*RespPK/(RespPK+C50))-kout*RespPD")

# 2. POPULATION PARAMETERS
modelParametersPKPD = list(
  ModelParameter( name = "V",    distribution = LogNormal( mu = 8, omega = sqrt(0.02) )),
  ModelParameter( name = "Cl",   distribution = LogNormal( mu = 0.13, omega = sqrt(0.06) ) ),
  ModelParameter( name = "ka",   distribution = LogNormal( mu = 1.6, omega = sqrt(0.70) ) ),
  
  ModelParameter( name = "Rin",    distribution = LogNormal( mu = 5.4, omega = sqrt(0.2) )),
  ModelParameter( name = "kout",   distribution = LogNormal( mu = 0.06, omega = sqrt(0.02) ) ),
  ModelParameter( name = "Imax",   distribution = LogNormal( mu = 1, omega = 0 ), fixedMu = TRUE, fixedOmega = TRUE ),
  ModelParameter( name = "C50",   distribution = LogNormal( mu = 1.2, omega = sqrt(0.01) ) )
  
)

# 3. ERROR MODEL
errorModelRespPK = Combined1( output = "RespPK", sigmaInter = 0.6, sigmaSlope = 0.07 )
errorModelRespPD = Constant( output = "RespPD", sigmaInter = 4 )
modelErrorPKPD = list( errorModelRespPK, errorModelRespPD )
```


```{r}
# 4. DESIGN DEFINITION

## sampling times  
# samplingTimesRespPD = SamplingTimes( outcome = "RespPD", samplings = c( 0, 1, 2, 6, 12, 24, 36, 48, 72, 96, 120, 144 ) )
samplingTimesRespPD = SamplingTimes( outcome = "RespPD", samplings = c( 0, 24, 36, 48, 72, 96, 120, 144 ) )

## arms definition
empiricalArmPKPD = Arm( name = "empiricalArm",
                        size = 32,
                        administrations = list( administrationRespPK ) ,
                        samplingTimes   = list( samplingTimesRespPK, samplingTimesRespPD ),
                        initialCondition = list( "RespPK" = 0, "RespPD" = 0 ) )

empiricalDesignPKPD = Design( name = "empiricalDesign", arms = list( empiricalArmPKPD ) )
```


```{r}
# FISHER INFORMATION MATRIX COMPUTATION
evaluationEmpiricalDesignFIMSettingsPKPD = Evaluation( name = "",
                                                       modelEquations = modelEquationsPKPD,
                                                       modelParameters = modelParametersPKPD,
                                                       modelError = modelErrorPKPD,
                                                       modelCovariates = list( CYP2C9 ),
                                                       modelCovariatesEquation = "exponential",
                                                       designs = list( empiricalDesignPKPD ),
                                                       outputs = list( "RespPK" = "RespPK", 
                                                                       "RespPD" = "RespPD" ),
                                                       fimType = 'population' ,
                                                       odeSolverParameters = list( atol = 1e-12, rtol = 1e-12  ) )

evaluationEmpiricalDesignFIMPKPD = run( evaluationEmpiricalDesignFIMSettingsPKPD )
show( evaluationEmpiricalDesignFIMPKPD )
```


# Q5 PK and PD sample optimisation
```{r}

samplingsArmOptimisationPK = SamplingTimes( outcome = "RespPK", samplings = c( 0.5, 6, 12, 96, 120) )
samplingsArmOptimisationPD = SamplingTimes( outcome = "RespPD", samplings = c( 0.5, 6, 12, 96, 120) )


administrationConstraintsMultRespPK = AdministrationConstraints( outcome = "RespPK", doses = list( 100 ) )



samplingConstraintsMultRespPD  = SamplingTimeConstraints( outcome = "RespPD",
                                                          initialSamplings = c( 0.5, 1, 2, 6, 12, 24, 48, 72, 96, 120 ),
                                                          numberOfsamplingsOptimisable = 5,
                                                          # fixedTimes = c()
)


armWithMultConstraintsPKPD = Arm( name = "armWithMultConstraints",
                                  size = 32,
                                  administrations = list( administrationRespPK ),
                                  administrationsConstraints = list( administrationConstraintsMultRespPK ),
                                  
                                  samplingTimes   = list( samplingsArmOptimisationPK, samplingsArmOptimisationPD ),
                                  samplingTimesConstraints = list( samplingConstraintsMultRespPD ),
                                  initialCondition = list( "RespPK" = 0, "RespPD" = 0 ) )


designWithMultConstraintsPKPD = Design( name = "designWithMultConstraints", arms = list( armWithMultConstraintsPKPD ), numberOfArms = 32 )

```


```{r}
# RUN MULTIPLICATIVE ALGORITHM OPTIMISATION
optimizationMultSettingsPKPD = Optimization( name = "Optimisation Mult",
                                             modelEquations = modelEquationsPKPD,
                                             modelParameters = modelParametersPKPD,
                                             modelError = modelErrorPKPD,
                                             modelCovariates = list( CYP2C9 ),
                                             modelCovariatesEquation = "exponential",
                                             optimizer = "MultiplicativeAlgorithm",
                                             optimizerParameters = list( lambda = 0.99,
                                                                         numberOfIterations = 1000,
                                                                         weightThreshold = 0.001,
                                                                         delta = 1e-04, showProcess = T ),
                                             designs = list( designWithMultConstraintsPKPD ),
                                             fimType = "population",
                                             outputs = list( "RespPK" = "RespPK", 
                                                             "RespPD" = "RespPD" ),
                                             odeSolverParameters = list( atol = 1e-12, rtol = 1e-12  ) )

optimizationMultFIMPKPD = run( optimizationMultSettingsPKPD )
show(optimizationMultFIMPKPD)

```


```{r}
show(optimizationMultFIM)
show(optimizationMultFIMPKPD)

```

